---
title: "gtm-introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gtm-introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

## What is GTM?

The Generative Topographic Mapping (GTM) is a probabilistic model for
nonlinear dimensionality reduction, introduced by Bishop, Svensen, and
Williams (1998). Like the Self-Organizing Map (SOM), GTM maps
high-dimensional data onto a low-dimensional grid for visualization.
Unlike SOM, GTM is grounded in a proper probability density model and is
trained by Expectation-Maximization (EM), which guarantees monotonic
improvement of the log-likelihood.

Key advantages over SOM:

- A principled probabilistic model (no ad-hoc learning rate or
  neighborhood function schedules)
- Quantifiable density estimates in data space
- Magnification factors from the Jacobian of the mapping
- Convergence is guaranteed and measurable

## Quick Start

Fit a GTM to the classic `iris` dataset:

```{r fit-iris}
library(gtm)
data(iris)

model <- GtmFit(as.matrix(iris[, 1:4]), n_iter = 50, verbose = FALSE)
print(model)
```

## Projection Plots

Project data points onto the 2D latent space using posterior mean
positions, colored by species:

```{r plot-means}
plot(model, type = "means", labels = iris$Species,
     main = "Iris species on GTM (mean projection)")
```

The mode projection assigns each point to its highest-responsibility node:

```{r plot-modes}
plot(model, type = "modes", labels = iris$Species,
     main = "Iris species on GTM (mode projection)")
```

## Exploring the Map

### Density

The density plot shows the cumulative responsibility at each node,
revealing which parts of the map are most populated:

```{r plot-density}
plot(model, type = "density")
```

### U-Matrix

The U-matrix shows the average distance between neighboring nodes in
data space. High values indicate cluster boundaries:

```{r plot-umatrix}
plot(model, type = "umatrix")
```

### Component Planes

Component planes show the R-weighted average of each data variable per
node. This reveals how each variable varies across the map:

```{r plot-components, fig.height = 6}
plot(model, type = "components")
```

Or inspect a single variable:

```{r plot-component-1}
plot(model, type = "components", component = 1,
     main = "Sepal.Length")
```

## Magnification Factors

A unique GTM feature: the magnification factor at each node measures
local area distortion of the mapping. Regions where the map is
stretched to cover more data space have higher magnification:

```{r plot-magnification}
plot(model, type = "magnification")
```

## Convergence

The log-likelihood should increase monotonically during EM training:

```{r plot-convergence}
plot(model, type = "convergence")
```

## Hexagonal Grid

GTM supports hexagonal grids, which provide more uniform neighbor
distances:

```{r hex-grid}
model_hex <- GtmFit(as.matrix(iris[, 1:4]),
                    grid_type = "hex", n_iter = 50)
plot(model_hex, type = "density", main = "Hex grid density")
```

## 1D Latent Space

GTM also works with a 1-dimensional latent space:

```{r fit-1d}
model_1d <- GtmFit(as.matrix(iris[, 1:4]),
                   latent_dim = 1, n_iter = 50)
print(model_1d)
plot(model_1d, type = "density", main = "1D GTM density")
```

## Prediction

Project new data onto a fitted map:

```{r predict}
new_data <- as.matrix(iris[1:5, 1:4])
proj_means <- predict(model, newdata = new_data, type = "means")
print(proj_means)
```

## References

Bishop, C. M., Svensen, M., & Williams, C. K. I. (1998).
GTM: The Generative Topographic Mapping.
*Neural Computation*, 10(1), 215-234.
